### Script for ChIPQC

```{r, load libraries, , include=FALSE}
## Load Libraries
library(BiocManager)
library(ChIPQC)
library(ggplot2)
```

```{r, set wd, include=FALSE}
#Set Working Directory
knitr::opts_chunk$set(echo = TRUE)

#Set working environment

workingdir="C:/Users/eddie/Research/GitHub"

#Set working directory to the correct location for working machine
knitr::opts_knit$set(root.dir = "workingdir")
```

```{r, Load Data & Run QC}
## Load Sample Data
samples <- read.csv('meta/DiffBind_CAF-1_ET.csv')

## Make ChIPQC Object
register(SerialParam())
chipObj <- ChIPQC(samples) 

## Make QC Report
ChIPQCreport(chipObj, reportName="ChIP_QC_report_cac_H3K27me3", reportFolder="ChIPQCreport")
```

### DiffBind (scratch)
```{r, DiffBind Analysis}
#Read in sample sheet containing CAF-1 mutant data *If running ChIP QC, you can use "samples" variable
caf_dba <- dba(sampleSheet = samples)

##Plot correlation plot between all samples included on data sheet, save in a file
CorrPlot <- plot(caf_dba)
ggsave(filename = "CAF-1_DBA_Corr_Plot.pdf", plot = CorrPlot, dpi=600)

#Count Reads
caf_dba <- dba.count(caf_dba)

##Plot new correlation plot based on count data, save in a file
CorrPlot_count <- plot(caf_dba)
ggsave(filename = "CAF-1_DBA_Corr_Plot_counts.pdf", plot = CorrPlot_count, dpi=600)

#Normalization
caf_dba <- dba.normalize(caf_dba)

#Model design & Contrast (whta comparisons do you want to make?)
caf_dba <- dba.contrast(caf_dba)

#Differential Analysis
caf_dba <- dba.analyze(caf_dba)

##Plot correlation plot based only on differentially bound sites, save in file
CorrPlot_diff <- plot(caf_dba, contrast = 1)
ggsave(filename = "CAF-1_DBA_Corr_Plot_diff.pdf", plot = CorrPlot_diff, dpi=600)

#Retrieve differentially bound sites for downstream analysis (will return a GRanges object)
caf_dba.DB <- dba.report(caf_dba)

```

```{r, DiffBind Plotting}

```

### DiffBind Script

## Load Libraries
library(DiffBind)
library(tidyverse)

## Run DBA
dbObj <- dba(sampleSheet=samples)

## Plot PCA & Correlation Plot
dba.plotPCA(dbObj,  attributes=DBA_FACTOR, label=DBA_ID)
plot(dbObj)

## Contrast Samples
dbObj <- dba.contrast(dbObj, categories=DBA_FACTOR, minMembers = 2)

## Differential Enrichment Analysis
dbObj <- dba.analyze(dbObj, method=DBA_ALL_METHODS)
dba.show(dbObj, bContrasts=T)

# Plot
dba.plotPCA(dbObj, contrast=1, method=DBA_DESEQ2, attributes=DBA_FACTOR, label=DBA_ID)

## Visualize Results

# Peaks recognized by edgeR vs. DEseq
dba.plotVenn(dbObj,contrast=1,method=DBA_ALL_METHODS)

# MA Plots
dba.plotMA(dbObj, contrast=1, method=DBA_DESEQ2)
dba.plotMA(dbObj, contrast=1, bXY=TRUE)
pvals <- dba.plotBox(dbObj, contrast=1)

## Extract Results
res_deseq <- dba.report(dbObj, method=DBA_DESEQ2, contrast = 3, th=1)

# Write to File
out <- as.data.frame(res_deseq)
write.table(out, file="cac3_H3K27me3_DiffBind.txt", sep="\t", quote=F, row.names=F)

# Create bed files for each keeping only significant peaks (p < 0.05)

WT_cac3_enrich <- out %>% 
  filter(FDR < 0.05 & Fold > 0) %>% 
  select(seqnames, start, end)

# Write to file
write.table(WT_cac3_enrich, file="WT_cac3_enriched.bed", sep="\t", quote=F, row.names=F, col.names=F)

cac3_enrich <- out %>% 
  filter(FDR < 0.05 & Fold < 0) %>% 
  select(seqnames, start, end)

# Write to file
write.table(cac3_enrich, file="cac3_enriched.bed", sep="\t", quote=F, row.names=F, col.names=F)

